

---
title: "Simulation Challenge"
subtitle: "Generative Models and Monte Carlo Simulation"
format:
  html: default
  pdf: default
execute:
  echo: true
  eval: true
---

# 🎲 Simulation Challenge - Monte Carlo Analysis

## Challenge Overview

**Your Mission:** Create a comprehensive Quarto document that simulates one or two investment strategies, analyzes the results, and demonstrates your ability to present counter-intuitive findings compellingly. Then render the document to HTML and deploy it via GitHub Pages from a new repository called "simulationChallenge."

## The Investment Game 🎯

### Original Game Strategy

::: {#exm-ErgodicityEconomicsExample}
Imagine you are offered the following game and given a $1,000 budget in a special account to play the game: I will flip a coin, and if it comes up heads, we increase your account's balance by 50%; if it comes up tails, we reduce your account's balance by 40%. We are not only doing this once, but we will do it once per year until you turn 55. When you turn 55, you will receive the balance in your account.
:::

### Generative DAG Model for the Investment Game

```{python}
#| echo: false
#| include: false
#| eval: false
import daft

# Create the DAG directly without custom class
investmentDAG = daft.PGM(dpi=150, alternate_style="outer")

# Initial wealth (deterministic)
investmentDAG.add_node("W0", "Initial Wealth\n$W_0 = 1000$", x=4, y=2.5, aspect=5.4, 
                      alternate=True, plot_params={'facecolor': 'aliceblue'})

# Time t nodes  
investmentDAG.add_node("Ct", "Coin Flip t\n$C_t \\sim \\text{Bernoulli}(0.5)$", x=0, y=1, aspect=4, scale = 1.6,
                      plot_params={'facecolor': 'aliceblue'})
investmentDAG.add_node("Wt", "\nWealth t\n$W_t = 1.5 \\times W_{t-1}$ if $C_t = 1$\n$W_t = 0.6 \\times W_{t-1}$ if $C_t = 0$\n", x=4, y=1, scale = 1.6, aspect=4,
                      alternate=True, plot_params={'facecolor': 'aliceblue'})

# Add edges showing the relationships
investmentDAG.add_edge("W0", "Wt")
investmentDAG.add_edge("Ct", "Wt")

# Add plate around time-dependent nodes
investmentDAG.add_plate([-2, 0.5, 7.8, 1.0], label="t = 1, ..., N", shift=-0.1)
```

```{python}
#| label: fig-investment-dag
#| fig-cap: Generative DAG model for the investment game showing how wealth evolves over time through coin flips
#| echo: false
#| eval: false
investmentDAG.show()
```

## Challenge Requirements 📋

### Questions to Answer for 75% Grade on Challenge

1. **Expected Value Analysis:** What is the "expected value" of your account balance after 1 coin flip for the original game?

2. **Expectation vs. Reality:** Is the expected value positive or negative? Do you expect your account to be worth more or less than $1,000 based on this result?

3. **Single Simulation:** Run one simulation showing the dynamics of your account balance over time. Make an object-oriented matplotlib OR ggplot2 plot showing your simulated account balance over time (i.e. as you age). Comment on the results, are you happy?

## Analysis and Solutions

### Question 1: Expected Value Analysis

**What is the "expected value" of your account balance after 1 coin flip for the original game?**

Let's calculate this step by step using R:

```{r}
#| label: expected-value-analysis
#| fig-cap: Expected value calculation for the investment game
#| echo: false

# Load required libraries
suppressPackageStartupMessages(library(tidyverse))

# Game parameters
initial_balance <- 1000
heads_multiplier <- 1.5  # +50% gain
tails_multiplier <- 0.6  # -40% loss
probability_heads <- 0.5

# Calculate expected value after 1 coin flip
# E[W1] = P(Heads) * W0 * 1.5 + P(Tails) * W0 * 0.6
# E[W1] = 0.5 * 1000 * 1.5 + 0.5 * 1000 * 0.6
# E[W1] = 0.5 * 1500 + 0.5 * 600
# E[W1] = 750 + 300 = 1050

expected_value_after_one_flip <- probability_heads * initial_balance * heads_multiplier + 
                                 (1 - probability_heads) * initial_balance * tails_multiplier

# Display the calculation
cat("Expected Value Calculation:\n")
cat("========================\n")
cat("Initial Balance: $", initial_balance, "\n")
cat("Heads (50% chance): $", initial_balance, " × 1.5 = $", initial_balance * heads_multiplier, "\n")
cat("Tails (50% chance): $", initial_balance, " × 0.6 = $", initial_balance * tails_multiplier, "\n")
cat("\nExpected Value = 0.5 × $", initial_balance * heads_multiplier, " + 0.5 × $", initial_balance * tails_multiplier, "\n")
cat("Expected Value = $", probability_heads * initial_balance * heads_multiplier, " + $", (1 - probability_heads) * initial_balance * tails_multiplier, "\n")
cat("Expected Value = $", expected_value_after_one_flip, "\n")

# Verify with simulation
set.seed(123)
n_sims <- 10000

# Simulate 10,000 single coin flips
simulated_balances <- replicate(n_sims, {
  coin_flip <- rbinom(1, 1, probability_heads)
  if (coin_flip == 1) {
    initial_balance * heads_multiplier
  } else {
    initial_balance * tails_multiplier
  }
})

# Calculate mean of simulations
simulated_expected_value <- mean(simulated_balances)

cat("\nSimulation Verification:\n")
cat("=======================\n")
cat("Simulated Expected Value (10,000 trials): $", round(simulated_expected_value, 2), "\n")
cat("Theoretical Expected Value: $", expected_value_after_one_flip, "\n")
cat("Difference: $", round(abs(simulated_expected_value - expected_value_after_one_flip), 2), "\n")
```

### Question 2: Expectation vs. Reality

**Is the expected value positive or negative? Do you expect your account to be worth more or less than $1,000 based on this result?**

```{r}
#| label: expectation-analysis
#| fig-cap: Analysis of expected value vs initial investment
#| echo: false

# Analysis of the expected value
cat("Expectation vs. Reality Analysis:\n")
cat("================================\n")
cat("Initial Investment: $", initial_balance, "\n")
cat("Expected Value after 1 flip: $", expected_value_after_one_flip, "\n")
cat("Expected Gain/Loss: $", expected_value_after_one_flip - initial_balance, "\n")
cat("Expected Return: ", round((expected_value_after_one_flip - initial_balance) / initial_balance * 100, 1), "%\n")

if (expected_value_after_one_flip > initial_balance) {
  cat("\n✅ The expected value is POSITIVE!\n")
  cat("Based on this result, I expect my account to be worth MORE than $1,000.\n")
  cat("This suggests the game is favorable in the long run.\n")
} else {
  cat("\n❌ The expected value is NEGATIVE!\n")
  cat("Based on this result, I expect my account to be worth LESS than $1,000.\n")
  cat("This suggests the game is unfavorable in the long run.\n")
}

# However, let's think about this more carefully...
cat("\nHowever, the expected value calculation assumes we can play this game many times.\n")
cat("But in reality, we only get to play once per year until age 55.\n")
```

### Question 3: Single Simulation

**Run one simulation showing the dynamics of your account balance over time. Make an object-oriented ggplot2 plot showing your simulated account balance over time (i.e. as you age). Comment on the results, are you happy?**

Let's simulate the investment game over multiple years to see how the account balance evolves:

```{r}
#| label: single-simulation
#| fig-cap: Single simulation of investment game showing account balance over time
#| echo: false

# Load required libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(456)

# Game parameters
initial_balance <- 1000
heads_multiplier <- 1.5  # +50% gain
tails_multiplier <- 0.6  # -40% loss
probability_heads <- 0.5
years_to_simulate <- 30  # Simulate from age 25 to 55

# Function to simulate one complete investment path
simulate_investment_path <- function(initial, years, heads_mult, tails_mult, prob_heads) {
  balance <- initial
  path <- numeric(years + 1)
  path[1] <- initial
  
  for (year in 1:years) {
    coin_flip <- rbinom(1, 1, prob_heads)
    if (coin_flip == 1) {
      balance <- balance * heads_mult  # Heads: multiply by 1.5
    } else {
      balance <- balance * tails_mult  # Tails: multiply by 0.6
    }
    path[year + 1] <- balance
  }
  
  return(path)
}

# Run single simulation
single_path <- simulate_investment_path(initial_balance, years_to_simulate, 
                                       heads_multiplier, tails_multiplier, probability_heads)

# Create data frame for plotting
simulation_data <- tibble(
  year = 0:years_to_simulate,
  age = 25 + year,  # Assuming starting at age 25
  balance = single_path,
  log_balance = log(balance)
)

# Display the simulation results
cat("Single Investment Simulation Results:\n")
cat("====================================\n")
cat("Starting Age: 25\n")
cat("Ending Age: 55\n")
cat("Initial Balance: $", initial_balance, "\n")
cat("Final Balance: $", round(single_path[length(single_path)], 2), "\n")
cat("Total Return: ", round((single_path[length(single_path)] / initial_balance - 1) * 100, 1), "%\n")
cat("Annualized Return: ", round((single_path[length(single_path)] / initial_balance)^(1/years_to_simulate) - 1, 3) * 100, "%\n")

# Show first few and last few years
cat("\nFirst 5 years:\n")
print(simulation_data[1:6, c("age", "balance")])

cat("\nLast 5 years:\n")
print(simulation_data[(nrow(simulation_data)-4):nrow(simulation_data), c("age", "balance")])

# Create the time series plot
p1 <- ggplot(simulation_data, aes(x = age, y = balance)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "darkblue", size = 2) +
  geom_hline(yintercept = initial_balance, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Investment Game: Single Simulation Path",
    subtitle = paste("Account Balance from Age 25 to 55"),
    x = "Age",
    y = "Account Balance ($)",
    caption = "Red dashed line shows initial investment ($1,000)"
  ) +
  scale_y_continuous(
    labels = scales::dollar_format(),
    trans = "log10",
    breaks = c(100, 1000, 10000, 100000)
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )

print(p1)

# Analysis of the results
cat("\n📊 Analysis of Single Simulation:\n")
cat("================================\n")

if (single_path[length(single_path)] > initial_balance) {
  cat("✅ This simulation was SUCCESSFUL!\n")
  cat("The account grew from $", initial_balance, " to $", round(single_path[length(single_path)], 2), "\n")
  cat("This represents a gain of $", round(single_path[length(single_path)] - initial_balance, 2), "\n")
} else {
  cat("❌ This simulation was UNSUCCESSFUL!\n")
  cat("The account shrank from $", initial_balance, " to $", round(single_path[length(single_path)], 2), "\n")
  cat("This represents a loss of $", round(initial_balance - single_path[length(single_path)], 2), "\n")
}

# Count wins vs losses by year
wins <- sum(single_path[-1] > single_path[-length(single_path)])
losses <- sum(single_path[-1] < single_path[-length(single_path)])

cat("\nYear-by-year breakdown:\n")
cat("Winning years: ", wins, "\n")
cat("Losing years: ", losses, "\n")
cat("Win rate: ", round(wins / (wins + losses) * 100, 1), "%\n")

# Key insights
cat("\n🤔 Key Insights:\n")
cat("===============\n")
cat("1. Despite a positive expected value, +5% per year, this single path shows the reality of randomness.\n")
cat("2. The multiplicative nature of the game means early losses are very hard to recover from.\n")
cat("3. But its hard to get a full understanding from this single simulation.\n")
cat("4. The log scale plot shows the exponential nature of the growth/decay process.\n")
```

### Question 4: Multiple Simulations

**Run 100 simulations modelling the dynamics of your account balance over time. Make an object-oriented ggplot2 plot showing a probability distribution of the 100 simulated account balance at age 55. Comment on the results, are you happy? Why or why not?**

Now let's run 100 simulations to see the distribution of outcomes:

```{r}
#| label: multiple-simulations
#| fig-cap: Distribution of final account balances from 100 simulations
#| echo: false

# Load required libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(789)

# Game parameters (same as before)
initial_balance <- 1000
heads_multiplier <- 1.5  # +50% gain
tails_multiplier <- 0.6  # -40% loss
probability_heads <- 0.5
years_to_simulate <- 30  # Simulate from age 25 to 55
n_simulations <- 100

# Function to simulate final balance (more efficient for many simulations)
simulate_final_balance <- function(initial, years, heads_mult, tails_mult, prob_heads) {
  balance <- initial
  for (year in 1:years) {
    coin_flip <- rbinom(1, 1, prob_heads)
    if (coin_flip == 1) {
      balance <- balance * heads_mult  # Heads: multiply by 1.5
    } else {
      balance <- balance * tails_mult  # Tails: multiply by 0.6
    }
  }
  return(balance)
}

# Run 100 simulations
cat("Running 100 simulations...\n")
final_balances <- replicate(n_simulations, 
  simulate_final_balance(initial_balance, years_to_simulate, 
                        heads_multiplier, tails_multiplier, probability_heads))

# Create data frame for analysis
simulation_results <- tibble(
  simulation_id = 1:n_simulations,
  final_balance = final_balances,
  log_final_balance = log(final_balances),
  above_initial = final_balances > initial_balance,
  return_pct = (final_balances / initial_balance - 1) * 100
)

# Summary statistics
summary_stats <- simulation_results %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    prob_above_initial = mean(above_initial),
    prob_above_10000 = mean(final_balance > 10000),
    prob_above_100000 = mean(final_balance > 100000)
  )

# Display summary statistics
cat("100 Simulations Summary:\n")
cat("=======================\n")
cat("Mean Final Balance: $", round(summary_stats$mean_balance, 2), "\n")
cat("Median Final Balance: $", round(summary_stats$median_balance, 2), "\n")
cat("Minimum Final Balance: $", round(summary_stats$min_balance, 2), "\n")
cat("Maximum Final Balance: $", round(summary_stats$max_balance, 2), "\n")
cat("Probability above $1,000: ", round(summary_stats$prob_above_initial * 100, 1), "%\n")
cat("Probability above $10,000: ", round(summary_stats$prob_above_10000 * 100, 1), "%\n")
cat("Probability above $100,000: ", round(summary_stats$prob_above_100000 * 100, 1), "%\n")

# Create histogram of final balances
p2 <- ggplot(simulation_results, aes(x = final_balance)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = initial_balance, color = "red", linetype = "dashed", linewidth = 1.5) +
  geom_vline(xintercept = summary_stats$mean_balance, color = "darkgreen", linetype = "solid", linewidth = 1.5) +
  labs(
    title = "Distribution of Final Account Balances",
    subtitle = paste("100 Simulations from Age 25 to 55"),
    x = "Final Balance ($)",
    y = "Frequency",
    caption = "Red dashed line: Initial investment ($1,000)\nGreen solid line: Mean final balance"
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    trans = "log10",
    breaks = c(100, 1000, 10000, 100000, 1000000)
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )

print(p2)

# Analysis of the results
cat("\n📊 Analysis of 100 Simulations:\n")
cat("===============================\n")

# Count successful vs unsuccessful simulations
successful_sims <- sum(simulation_results$above_initial)
unsuccessful_sims <- n_simulations - successful_sims

cat("Successful simulations (>$1,000): ", successful_sims, " (", round(successful_sims/n_simulations*100, 1), "%)\n")
cat("Unsuccessful simulations (<$1,000): ", unsuccessful_sims, " (", round(unsuccessful_sims/n_simulations*100, 1), "%)\n")

# Extreme outcomes
extreme_wins <- sum(simulation_results$final_balance > 100000)
extreme_losses <- sum(simulation_results$final_balance < 100)

cat("\nExtreme outcomes:\n")
cat("Mega wins (>$100,000): ", extreme_wins, " simulations\n")
cat("Total losses (<$100): ", extreme_losses, " simulations\n")

# Key insights
cat("\n🤔 Key Insights from 100 Simulations:\n")
cat("====================================\n")
cat("1. The results are really skewed - most people lose money, but a few get very rich.\n")
cat("2. Even though the expected value is positive, the median outcome is much lower than the mean.\n")
cat("3. This shows how ensemble averages don't equal time averages - classic ergodicity breaking.\n")
cat("4. The multiplicative nature creates extreme inequality in outcomes.\n")
cat("5. Most people would be unhappy with this investment strategy!\n")

# Answer the question directly
cat("\n❓ Am I happy with these results?\n")
cat("===============================\n")
if (summary_stats$prob_above_initial > 0.5) {
  cat("✅ YES - More than 50% of simulations were successful.\n")
} else {
  cat("❌ NO - Less than 50% of simulations were successful.\n")
}

cat("The probability of ending up with more than $1,000 is only ", 
    round(summary_stats$prob_above_initial * 100, 1), "%.\n")
cat("This means there's a ", round((1 - summary_stats$prob_above_initial) * 100, 1), 
    "% chance of losing money over 30 years!\n")

cat("\nThis is a perfect example of why expected value can be misleading\n")
cat("when you only get to play the game once (or a limited number of times).\n")
```

### Question 5: Probability Analysis

**Based on the 100 simulations above, what is the probability that your account balance will be greater than $1,000 at age 55?**

Let's analyze the probability results from our 100 simulations:

```{r}
#| label: probability-analysis
#| fig-cap: Probability analysis of final account balances
#| echo: false

# Load required libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# We'll use the same simulation results from Question 4
# (In a real document, we'd reference the previous results, but for clarity we'll recalculate)

# Set seed for reproducibility (same as Question 4)
set.seed(789)

# Game parameters
initial_balance <- 1000
heads_multiplier <- 1.5
tails_multiplier <- 0.6
probability_heads <- 0.5
years_to_simulate <- 30
n_simulations <- 100

# Function to simulate final balance
simulate_final_balance <- function(initial, years, heads_mult, tails_mult, prob_heads) {
  balance <- initial
  for (year in 1:years) {
    coin_flip <- rbinom(1, 1, prob_heads)
    if (coin_flip == 1) {
      balance <- balance * heads_mult
    } else {
      balance <- balance * tails_mult
    }
  }
  return(balance)
}

# Run 100 simulations
final_balances <- replicate(n_simulations, 
  simulate_final_balance(initial_balance, years_to_simulate, 
                        heads_multiplier, tails_multiplier, probability_heads))

# Calculate key probabilities
prob_above_1000 <- mean(final_balances > 1000)
prob_above_10000 <- mean(final_balances > 10000)
prob_above_100000 <- mean(final_balances > 100000)
prob_below_1000 <- mean(final_balances < 1000)
prob_below_100 <- mean(final_balances < 100)

# Calculate confidence interval for the probability
# Using normal approximation for binomial proportion
n <- n_simulations
p_hat <- prob_above_1000
se <- sqrt(p_hat * (1 - p_hat) / n)
z_score <- 1.96  # 95% confidence interval
ci_lower <- p_hat - z_score * se
ci_upper <- p_hat + z_score * se

# Display probability analysis
cat("Probability Analysis Results:\n")
cat("============================\n")
cat("Based on 100 simulations:\n\n")

cat("🎯 MAIN QUESTION: Probability of balance > $1,000 at age 55:\n")
cat("   Answer: ", round(prob_above_1000 * 100, 1), "%\n")
cat("   95% Confidence Interval: [", round(ci_lower * 100, 1), "%, ", round(ci_upper * 100, 1), "%]\n\n")

cat("📊 Additional Probability Breakdown:\n")
cat("Probability > $1,000:  ", round(prob_above_1000 * 100, 1), "% (", sum(final_balances > 1000), " out of 100)\n")
cat("Probability > $10,000: ", round(prob_above_10000 * 100, 1), "% (", sum(final_balances > 10000), " out of 100)\n")
cat("Probability > $100,000:", round(prob_above_100000 * 100, 1), "% (", sum(final_balances > 100000), " out of 100)\n")
cat("Probability < $1,000:  ", round(prob_below_1000 * 100, 1), "% (", sum(final_balances < 1000), " out of 100)\n")
cat("Probability < $100:    ", round(prob_below_100 * 100, 1), "% (", sum(final_balances < 100), " out of 100)\n")

# Create a probability summary plot
prob_data <- tibble(
  threshold = c("$100", "$1,000", "$10,000", "$100,000"),
  probability = c(prob_below_100, prob_below_1000, 1 - prob_above_10000, 1 - prob_above_100000),
  category = c("Total Loss", "Below Initial", "Modest Gain", "Mega Loss")
) %>%
  mutate(
    threshold_value = c(100, 1000, 10000, 100000),
    probability_pct = round(probability * 100, 1)
  )

# Create probability bar chart
p3 <- ggplot(prob_data, aes(x = reorder(threshold, threshold_value), y = probability_pct)) +
  geom_col(fill = "steelblue", alpha = 0.7, color = "white") +
  geom_text(aes(label = paste0(probability_pct, "%")), 
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(
    title = "Probability of Different Outcomes",
    subtitle = "Based on 100 Simulations of Investment Game",
    x = "Outcome Threshold",
    y = "Probability (%)",
    caption = "Shows probability of NOT exceeding each threshold"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )

print(p3)

# Key insights
cat("\n🤔 Key Insights from Probability Analysis:\n")
cat("==========================================\n")
cat("1. The probability of ending up with MORE than $1,000 is only ", round(prob_above_1000 * 100, 1), "%\n")
cat("2. This means there's a ", round(prob_below_1000 * 100, 1), "% chance of LOSING money over 30 years!\n")
cat("3. Even though the expected value is positive (+5% per year), most people lose money.\n")
cat("4. Only ", round(prob_above_100000 * 100, 1), "% of people become mega-rich (>$100,000).\n")
cat("5. This shows how ergodicity breaking works in economics.\n")

# Theoretical vs empirical comparison
theoretical_prob <- 0.5  # This would be the theoretical probability if the game were fair
empirical_prob <- prob_above_1000

cat("\n📈 Theoretical vs Empirical Comparison:\n")
cat("=====================================\n")
cat("Theoretical probability (if game were fair): 50%\n")
cat("Empirical probability (from 100 sims): ", round(empirical_prob * 100, 1), "%\n")
cat("Difference: ", round((empirical_prob - theoretical_prob) * 100, 1), " percentage points\n")

if (empirical_prob < 0.5) {
  cat("\n❌ The empirical probability is LOWER than 50%!\n")
  cat("This confirms that the investment game is unfavorable for most players.\n")
} else {
  cat("\n✅ The empirical probability is HIGHER than 50%!\n")
  cat("This would suggest the game is favorable for most players.\n")
}

cat("\n🎯 FINAL ANSWER TO QUESTION 5:\n")
cat("==============================\n")
cat("The probability that your account balance will be greater than $1,000 at age 55\n")
cat("is approximately ", round(prob_above_1000 * 100, 1), "% based on 100 simulations.\n")
cat("This means you have a ", round(prob_below_1000 * 100, 1), "% chance of losing money!\n")
```

### Question 6: Strategy Comparison

**Run 100 simulations for the modified game strategy shown below. What is the probability that your account balance will be greater than $10,000 at age 55? Is this probability higher or lower than the probability in the original game?**

First, let's define the modified strategy:

#### Modified Game Strategy

::: {#exm-ErgodicityEconomicsExampleModified}
Imagine you are offered the following game and given a $1,000 budget in a special account to play the game: I will flip a coin, and if it comes up heads, we increase your bet by 50%; if it comes up tails, we reduce your bet by 40%. You must bet exactly 50% of your current account balance on each flip, and this 50% is locked in for each round. We are not only doing this once, but we will do it once per year until you turn 55. When you turn 55, you will receive the balance in your account.
:::

**Key Difference:** In the modified strategy, you only bet 50% of your balance each round, so you can't lose everything in one bad flip.

Now let's compare both strategies:

```{r}
#| label: strategy-comparison
#| fig-cap: Comparison of original vs modified investment strategies
#| echo: false

# Load required libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(999)

# Game parameters
initial_balance <- 1000
years_to_simulate <- 30
n_simulations <- 100

# ORIGINAL STRATEGY: Bet entire balance each round
# Heads: multiply by 1.5, Tails: multiply by 0.6
simulate_original_strategy <- function(initial, years) {
  balance <- initial
  for (year in 1:years) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance * 1.5  # +50% on entire balance
    } else {
      balance <- balance * 0.6  # -40% on entire balance
    }
  }
  return(balance)
}

# MODIFIED STRATEGY: Bet only 50% of balance each round
# Heads: +50% on the bet, Tails: -40% on the bet
simulate_modified_strategy <- function(initial, years) {
  balance <- initial
  for (year in 1:years) {
    coin_flip <- rbinom(1, 1, 0.5)
    bet_amount <- balance * 0.5  # Bet 50% of current balance
    
    if (coin_flip == 1) {
      # Heads: win 50% on the bet
      balance <- balance - bet_amount + bet_amount * 1.5
    } else {
      # Tails: lose 40% on the bet
      balance <- balance - bet_amount + bet_amount * 0.6
    }
  }
  return(balance)
}

# Run simulations for both strategies
cat("Running 100 simulations for each strategy...\n")

# Original strategy simulations
original_balances <- replicate(n_simulations, 
  simulate_original_strategy(initial_balance, years_to_simulate))

# Modified strategy simulations
modified_balances <- replicate(n_simulations, 
  simulate_modified_strategy(initial_balance, years_to_simulate))

# Create comparison data frame
comparison_data <- tibble(
  simulation_id = rep(1:n_simulations, 2),
  strategy = rep(c("Original", "Modified"), each = n_simulations),
  final_balance = c(original_balances, modified_balances),
  above_1000 = c(original_balances > 1000, modified_balances > 1000),
  above_10000 = c(original_balances > 10000, modified_balances > 10000),
  above_100000 = c(original_balances > 100000, modified_balances > 100000)
)

# Calculate probabilities for each strategy
strategy_summary <- comparison_data %>%
  group_by(strategy) %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    prob_above_1000 = mean(above_1000),
    prob_above_10000 = mean(above_10000),
    prob_above_100000 = mean(above_100000),
    prob_below_1000 = mean(!above_1000)
  )

# Display comparison results
cat("Strategy Comparison Results:\n")
cat("===========================\n\n")

cat("📊 SUMMARY STATISTICS:\n")
cat("======================\n")
print(strategy_summary)

cat("\n🎯 KEY PROBABILITIES:\n")
cat("=====================\n")
cat("Original Strategy:\n")
cat("  P(balance > $1,000):  ", round(strategy_summary$prob_above_1000[1] * 100, 1), "%\n")
cat("  P(balance > $10,000): ", round(strategy_summary$prob_above_10000[1] * 100, 1), "%\n")
cat("  P(balance > $100,000):", round(strategy_summary$prob_above_100000[1] * 100, 1), "%\n\n")

cat("Modified Strategy:\n")
cat("  P(balance > $1,000):  ", round(strategy_summary$prob_above_1000[2] * 100, 1), "%\n")
cat("  P(balance > $10,000): ", round(strategy_summary$prob_above_10000[2] * 100, 1), "%\n")
cat("  P(balance > $100,000):", round(strategy_summary$prob_above_100000[2] * 100, 1), "%\n")

# Create comparison plots
p4 <- ggplot(comparison_data, aes(x = final_balance, fill = strategy)) +
  geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 1000, color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 10000, color = "darkgreen", linetype = "dashed", linewidth = 1) +
  facet_wrap(~strategy, ncol = 2) +
  labs(
    title = "Distribution of Final Balances: Strategy Comparison",
    subtitle = "100 Simulations Each - Original vs Modified Strategy",
    x = "Final Balance ($)",
    y = "Frequency",
    caption = "Red line: $1,000 | Green line: $10,000"
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    trans = "log10",
    breaks = c(100, 1000, 10000, 100000, 1000000)
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    legend.position = "none"
  )

print(p4)

# Direct comparison of key probabilities
prob_original_10k <- strategy_summary$prob_above_10000[1]
prob_modified_10k <- strategy_summary$prob_above_10000[2]

cat("\n🎯 ANSWER TO QUESTION 6:\n")
cat("========================\n")
cat("Probability of balance > $10,000 at age 55:\n")
cat("  Original Strategy:  ", round(prob_original_10k * 100, 1), "%\n")
cat("  Modified Strategy:  ", round(prob_modified_10k * 100, 1), "%\n")

if (prob_modified_10k > prob_original_10k) {
  cat("\n✅ The MODIFIED strategy has a HIGHER probability!\n")
  cat("Difference: +", round((prob_modified_10k - prob_original_10k) * 100, 1), " percentage points\n")
} else {
  cat("\n❌ The ORIGINAL strategy has a HIGHER probability!\n")
  cat("Difference: +", round((prob_original_10k - prob_modified_10k) * 100, 1), " percentage points\n")
}

# Key insights
cat("\n🤔 Key Insights from Strategy Comparison:\n")
cat("=========================================\n")
cat("1. The modified strategy is more conservative - you can't lose everything in one flip.\n")
cat("2. This reduces extreme losses but also limits extreme gains.\n")
cat("3. The modified strategy should have a higher probability of modest success.\n")
cat("4. However, it may have a lower probability of mega-wins.\n")
cat("5. This shows the risk-return tradeoff in investment strategies.\n")

# Risk analysis
cat("\n📈 RISK ANALYSIS:\n")
cat("=================\n")
cat("Original Strategy Risk:\n")
cat("  - Can lose everything in one bad flip\n")
cat("  - Higher volatility (extreme outcomes)\n")
cat("  - Higher potential for both gains and losses\n\n")

cat("Modified Strategy Risk:\n")
cat("  - Can only lose 50% of balance per flip\n")
cat("  - Lower volatility (more stable outcomes)\n")
cat("  - More predictable, less extreme results\n")

cat("\n🎯 FINAL CONCLUSION:\n")
cat("====================\n")
cat("The modified strategy is generally better for most people because:\n")
cat("1. It reduces the risk of total loss\n")
cat("2. It provides more consistent, predictable outcomes\n")
cat("3. It has a higher probability of modest success\n")
cat("4. It's less dependent on getting lucky early\n")
cat("\nThis is a perfect example of why diversification and risk management\n")
cat("are crucial in real-world investing!\n")
```

## Project Analysis and Review

This simulation challenge provided a fascinating exploration into the world of generative models and Monte Carlo simulation through the lens of an investment game. By modeling a simple coin-flipping investment strategy, we uncovered profound insights about probability, risk, and the often counterintuitive nature of financial decision-making. The project systematically built from basic expected value calculations to complex multi-strategy comparisons, revealing how mathematical theory can diverge dramatically from real-world outcomes. Through 100 simulations of each strategy, we demonstrated the power of computational methods to illuminate patterns that would be impossible to see through analytical methods alone.

One thing I learned from this project is the concept of ergodicity breaking, where ensemble averages differ from time averages. Despite a positive expected value of +5% per year, the majority of individual investors would actually lose money over the 30-year period. 

Furthermore, it is Interesting to compare the original strategy, betting entire balance, and the modified strategy, betting only 50% of balance. This shows important insights about risk management, with the modified approach providing more consistent outcomes and higher probabilities of modest success.


